Two Decades of Smalltalk VM Development

### 2018-08-16 paper.tex
Minor tweaks (more keywords, cite Monticello, Spur for perf and 64-bits).

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

%% Keywords
%% comma separated list
\keywords{Just-in-Time compiler, virtual machine, managed runtime, tools}  %% \keywords are mandatory in final camera-ready submission

@@

To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built Squeak \cite{BackToTheFuture}, an open-source Smalltalk dialect, Squeak \cite{SqueakByExample}, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from \cite{blueBook} as a starting point. Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.  Development was done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining the generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.

@@

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. 
As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
%The code base was liberally sprinkled with assertions. 
The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  
Finally, a new object representation and garbage collector was added to also support 64-bits \cite{Spur}, which required refactoring the interpreter to allow the object representation to be chosen at startup. Bochs was used a second time for x64 support.

@@

The VM can actually be compiled in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is used in production, the interpreter version is convenient for development purposes. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding the JIT complexity.

@@

When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.
\egb{there is overlap here with respect to the previous paragraph. We already knew that a snapshot can be made during a coding session, and what a snapshot includes}
  When the programmer chooses, he can take a new snapshot, which includes his code changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS]. \egb{deployment here refers to starting a Smalltalk system from the previous paragraph, right?}Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello citation?  wiresong.ca}

>>>>

\usepackage{xcolor}
\newcommand{\todo}[1]{\color{orange}\fbox{\bfseries\sffamily\scriptsize TODO:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\sd}[1]{\color{red}\fbox{\bfseries\sffamily\scriptsize Stef:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\egb}[1]{\color{blue}\fbox{\bfseries\sffamily\scriptsize Elisa:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\eem}[1]{\color{green}\fbox{\bfseries\sffamily\scriptsize Eliot:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
\newcommand{\cba}[1]{\color{purple}\fbox{\bfseries\sffamily\scriptsize Clement:}{\sf\small$\blacktriangleright$\textit{#1}$\blacktriangleleft$}\color{black}}
%\newcommand*{rotatebox{75}}
\input{macros}

@@

%% Keywords
%% comma separated list
\keywords{Just-in-Time compiler, garbage collector, virtual machine, managed runtime, tools, live development}  %% \keywords are mandatory in final camera-ready submission

@@

To specify the virtual machine (VM), the Smalltalk-80 team at Xerox PARC wrote a Smalltalk VM entirely in Smalltalk \cite{blueBook}. In 1995, members of the same team built Squeak \cite{SqueakByExample}, an open-source Smalltalk dialect, and its VM \cite{BackToTheFuture}, written in Smalltalk using the code from \cite{blueBook} as a starting point. Part of the code base was, however, narrowed down to a subset of Smalltalk, called \emph{Slang}, to allow Smalltalk to C compilation.  Development was done in Smalltalk, an activity we call "Simulation", and the support code for which is "The Simulator". The production VM is derived by translating the core VM code to C, combining this generated C code with a set of platform-specific support files, and compiling with the platform's C compiler.

@@

Over the years, the Squeak VM evolved to give birth recently to \openSmalltalkVM\footnote{https://github.com/OpenSmalltalk/opensmalltalk-vm/}, the default VM for various Smalltalk-like systems such as Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample}, Cuis, Croquet and Newspeak \cite{NewspeakOopsla}. 
As the VM evolved, the original simulator co-evolved as a tool to develop and debug the VM. The meta data used to guide the translation process was replaced by \emph{pragmas} \cite{Pragmas}.  
%One change moved the meta data that guides translation out of the executable Smalltalk into pragmas \cite{Pragmas}.  
But the most significant evolution of the simulator came with the introduction of the Just-In-Time compiler (JIT). 
%The code base was liberally sprinkled with assertions. 
The spaghetti stack was mapped to a more conventional stack frame organisation in a stack zone of a few hundred k bytes organised into small pages, a scheme called context-to-stack mapping \cite{EfficientST80}, \cite{StackMappingVW}.  Machine code generated by the JIT was executed by binding multiple processor simulators (first Bochs \cite{Bochs} for x86, SkyEye for ARMv6, Smalltalk code for 32-bit MIPS).  
Finally, a new object representation and garbage collector was added to improve performance and also to support 64-bits \cite{Spur}.  This required refactoring the interpreter to allow the object representation to be chosen at startup \footnote{The JIT was written with this eventuality in mind.}. Bochs was used a second time for x64 support.

@@

The VM can actually be configured in two main flavours, interpreter-only or interpreter+JIT. Although the version with the JIT is most often used in production, the interpreter version is convenient for development purposes \footnote{and on devices that outlaw JITs}. For example, debugging the garbage collector or evaluating new language features can be done in the interpreter-only VM, avoiding the JIT complexity.

@@

When programming with Smalltalk, the programmer usually starts from a snapshot which contains the core libraries, the development environment and the application under development. Development of the application consists essentially in writing and editing code, which effectively installs, modifies and removes classes and compiled methods from the set of live objects. Programming may be done live, as the application under development is running, for example creating or revising methods in the debugger, and continuing.  Objects may have their shape changed on the fly as instance variables are added and removed.
\egb{there is overlap here with respect to the previous paragraph. We already knew that a snapshot can be made during a coding session, and what a snapshot includes}
  When the programmer chooses, he can take a new snapshot, which includes his code changes, and may include application instances under development [MAYBE THIS SHOULD BE DESCRIBED BELOW IN SNAPSHOTS]. \egb{deployment here refers to starting a Smalltalk system from the previous paragraph, right?}Deployment is performed using a snapshot containing the deployed application and the required core libraries (unused libraries and development tools may be removed from the deployed snapshot to decrease memory footprint and for security purposes).  \footnote{Code is organised in packages which are saved to an external version control system, in our case Monticello \cite{Monticello}}.
